Verifying Spatial Properties of Array Computations

### 2017-04-18 oopsla17/model.tex
typos

\noindent
We define a lattice model of array access patterns which serves as
a semantic model of our specification language and a model of array
access patterns in source code. This model (1) used to explain the
meaning of our specifications; (2) as a basis for the inference and
checking algorithms (\Cref{sec:algorithms}); (3) justifies an equational
theory for specifications in the next section; (4) is used to optimise
specifications using lattice identities; and (5) can be used to guide
correct implementations.

The model is defined over vectors of sets of integers which we call
\emph{index schemes}.  As an initial informal example, consider the
a stencil computation with loop body \fortran{y(i) = x(i,0) + x(i+1,0)}.
The access pattern on array \term{x}, relative
to induction variable \term{i}, is captured
by a vector of length 2 containing integer sets $\langle{\{0, 1\},
  \mathbb{Z}\}}\rangle$. This describes that, in the first dimension,
the array is read at offsets of $0$ and $1$ from an induction
variable. In the second dimension, the index is unconstrained
as it is a constant.

@@

\begin{defn}[Index scheme]
  An $N$-dimensional \emph{index scheme} is a vector of length $N$ of
  integer sets, \ie{}, a member of $\mathcal{P}(\bz)^{N}$.
  An equivalent view of these vectors is as an $N$-times Cartesian
  product on subsets of $\bz$.
  We use $S, T, U$ to denote index schemes. Henceforth, we assume index schemes
  are all $N$-dimensional\footnote{The particular
    dimensionality is derived from array types when the model is
    used.} for some $N$.

@@

\begin{restatable}{lem}{vectorUnion}
\label{lem:vector-union} If
$S$ and $T$ are index schemes where $\pi_i(S) = \pi_i(T)$ for all $1 \leq i
\leq N$ apart from some dimension $k$, then:
%
  \begin{equation*}
    S \cup  T
    =
    \pi_1(S) \times \cdots \times
    (\pi_k(S) \cup \pi_k(T)) \times \cdots \times
    \pi_N(S)
  \end{equation*}
\end{restatable}
%
\begin{defn}[Intervals with an optional hole]
  We define an extended notion of closed interval on $\bz{}$ which may contain a
  \emph{hole} at the origin, written \interv{a}{b}{c} where $a$ and $b$ are
  drawn from \bz{} with $a \leq 0 \leq b$ and $c$ is drawn from $\mathbb{B} = \{
  \mathsf{true}, \mathsf{false} \}$. Intervals are interpreted as sets as
  follows:
%
  \begin{equation*}
    \interv{a}{b}{c} \triangleq
      \{ n \mid a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
  \end{equation*}
  %
  If the
  superscript to the interval is omitted it is treated as $\mathsf{true}$.
  We also add the distinguished interval $\interv{-\infty}{\infty}{}$, which is simply an
  alias for \bz{}, but this notation prevents handling infinite interval
  separately in the following definitions, lemmas, and proofs. Here, $-\infty$
  and $\infty$ behave as top and bottom elements to \bz{} respectively.

@@

We define two specialisations of index schemes:
\emph{subscript scheme} and \emph{interval scheme}.

@@

\begin{defn}[Subscript scheme]
  A \emph{subscript scheme} is an index scheme scheme where:
  %
  \begin{equation*}
    \forall i.\ 1 \leq i \leq N \implies
      \pi_i (S) = \{ p \}
      \; \vee \;
      \pi_i (S) = \interv{-\infty}{\infty}{}
    \end{equation*}
%
  That is, the $i^{th}$ component of the set is either a singleton in \bz{} or
  the infinite interval.
\end{defn}
%
\begin{defn}[Region]
  A region is an index scheme and \region{N} is the set of all regions (\ie{},
  $\region{N} \subseteq \bz^N)$. The set of all regions is  defined
  as the smallest set satisfying the following:
%
  \begin{enumerate}
    \item If $R$ is in $\textit{Interval}^N$, then $R$ is in \region{N}.
    \item If $R$ and $S$ are in \region{N}, then so are $R \cap S$ and
      $R \cup S$.
  \end{enumerate}
\end{defn}
%
\begin{restatable}{prop}{regionLattice}
  \label{prop:region-lattice}
  $(\region{N},\cup,\cap,\subseteq)$ is a bounded distributive lattice with top
  $\top = \bz{}^N$ and bottom $\bot = \emptyset$.
\end{restatable}
%
The set of regions $\region{N}$ is target of our specification
language model, and of the code analysis. We further wrap this in a
labelled variant which provides information on multiplicity and approximation.

@@

We mark in our model the presence of modifiers such as
\texttt{readOnce} and \texttt{atMost}.  Approximation modifiers are interpreted as
injections into the $\mathsf{Approx}$ variant by $\interpApprox{-}$ in
\Cref{subfig:modifier-model}.  The $\textsf{Approx}$ type corresponds
to the presence or absence of the spatial approximation modifier, with
\textsf{exact} when there is no such modifier and \textsf{lower} and
\textsf{upper} for \term{atLeast} and \term{atMost}. In a similar way,
multiplicity modifiers are interpreted as injections in the
$\mathsf{Mult}$ variant by $\interpMult{-}$, corresponding to
the presence or absence of the \term{readOnce} modifier as shown in
\Cref{subfig:modifier-model}.

@@

\begin{defn}
  \label{defn:subscript}
  Recall array subscript terms of the form $a(\bar{e})$ from
  \Cref{def:array-subs}. We interpret these terms with the partial
  interpretation $\interp{-}^{\mathit{aterm}} : \textit{array-term}
  \prightarrow{} \mathcal{P}(\bz{})^N$. The interpretation is defined when
  all indices are either constant or neighbourhood indices as defined in
  \Cref{def:neighbour-ix}.
%
  \begin{align*}
    \interp{a(\bar{e})}^{\mathit{aterm}} =
      \prod_{1 \leq i \leq N} \mathit{subscript}(\bar{e}_i) & &
  %
    \textit{subscript}(e) = \begin{cases}
      \{ c \} & e \equiv i \pm c \\
      \bz & e \; \mbox{is absolute}
    \end{cases}
  \end{align*}
\end{defn}
%
Note that this produce subscript schemes (a form of index scheme), but
they are not members of \textit{Interval} or $\textit{Region}_N$. In
the next section, we use this interpretation on array subscripts in
the static analysis producer preceding checking and inference.

@@

Recall \Cref{lem:vector-intersect} states that the
intersection of index schemes is an index scheme. When
combined with \Cref{lem:interval-identities} (union and intersection
on holed intervals), we can extend this closure to \emph{interval} schemes.
Subscripts schemes do not enjoy the same closure propery as the intersection of
two singleton sets may be empty, disqualifying the result from 
being a subscript index. In the following sections, we do not take
intersection of subscript indices.

Union is not as flexible as
intersection. \Cref{lem:vector-union} does not form a closure for
indexing schemes in general. For this reason, we represent sets of
indexing schemes as unevaluated union terms, or \emph{union normal
  form}.  Since the model forms a distributive lattice as established
in \Cref{prop:region-lattice}, the intersections can be pushed inwards
using the distributive law leaving only unions at the outer-level. We
further exploit associativity of union to put unions into a cons-tree
making the unions effectively a non-empty list.  We do not, however,
attempt to canonicalise the representation. Later on
\Cref{lem:vector-union} and index scheme subset are used to reduce the
size of union normal form.

>>>>

\noindent
We define a lattice model of array access patterns which serves as
a semantic model of our specification language and a model of array
access patterns in source code. This model (1) is used to explain the
meaning of our specifications; (2) provides a basis for the inference and
checking algorithms (\Cref{sec:algorithms}); (3) justifies an equational
theory for specifications in the next section; (4) is used to optimise
specifications using lattice identities; and (5) can be used to guide
correct implementations.

The model is defined over vectors of sets of integers which we call
\emph{index schemes}.  As an initial informal example, consider the
 stencil kernel \fortran{y(i) = x(i,0) + x(i+1,0)}.
The access pattern on array \term{x}, relative
to induction variable \term{i}, is captured
by a vector containing two integer sets: $\langle{\{0, 1\},
  \mathbb{Z}\}}\rangle$. This describes that, in the first dimension,
the array is read at offsets of $0$ and $1$ from an induction
variable. In the second dimension, the index is unconstrained
as it is a constant.

@@

\begin{defn}[Index scheme]
  An $N$-dimensional \emph{index scheme} is a vector of containing $N$
  integer sets, \ie{}, a member of $\mathcal{P}(\bz)^{N}$.
  An equivalent view of these vectors is as an $N$-times Cartesian
  product on subsets of $\bz$.
  We use $S, T, U$ to denote index schemes. Henceforth, we assume index schemes
  are all $N$-dimensional\footnote{The particular
    dimensionality is derived from array types when the model is
    used.} for some $N$.

@@

\begin{restatable}{lem}{vectorUnion}
\label{lem:vector-union} If
$S$ and $T$ are index schemes where $\pi_i(S) = \pi_i(T)$ for all $1 \leq i
\leq N$ apart from some dimension $k$, then:
%
  \begin{equation*}
    S \cup  T
    =
    \pi_1(S) \times \cdots \times
    (\pi_k(S) \cup \pi_k(T)) \times \cdots \times
    \pi_N(S)
  \end{equation*}
\end{restatable}
%
\begin{defn}[Intervals with an optional hole]
  We define an extended notion of closed interval on $\bz{}$ which may contain a
  \emph{hole} at the origin, written \interv{a}{b}{c} where $a$ and $b$ are
  drawn from \bz{} with $a \leq 0 \leq b$ and $c$ is drawn from $\mathbb{B} = \{
  \mathsf{true}, \mathsf{false} \}$. Intervals are interpreted as sets as
  follows:
%
  \begin{equation*}
    \interv{a}{b}{c} \triangleq
      \{ n \mid a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
  \end{equation*}
  %
  If the
  superscript to the interval is omitted it is treated as $\mathsf{true}$ (no hole).
  We also add the distinguished interval $\interv{-\infty}{\infty}{}$, which is simply an
  alias for \bz{}, but this notation avoids separate handling of the infinite interval
  in the following definitions, lemmas, and proofs. Here, $-\infty$
  and $\infty$ behave as top and bottom elements to \bz{} respectively.

@@

We define two specialisations of index schemes:
the \emph{subscript scheme} and the \emph{interval scheme}.

@@

\begin{defn}[Subscript scheme]
  A \emph{subscript scheme} is an index scheme scheme where:
  %
  \begin{equation*}
    \forall i.\ 1 \leq i \leq N \implies
      \pi_i (S) = \{ p \}
      \; \vee \;
      \pi_i (S) = \interv{-\infty}{\infty}{}
    \end{equation*}
%
  That is, the $i^{th}$ component of the set is either a singleton in \bz{} or
  the infinite interval.
\end{defn}
%
\begin{defn}[Region]
  A region is an index scheme and \region{N} is the set of all regions (\ie{},
  $\region{N} \subseteq \bz^N)$. The set of all regions is  defined
  as the smallest set satisfying the following:
%
  \begin{enumerate}
    \item If $R$ is in $\textit{Interval}^N$, then $R$ is in \region{N}.
    \item If $R$ and $S$ are in \region{N}, then so are $R \cap S$ and
      $R \cup S$.
  \end{enumerate}
\end{defn}
%
\begin{restatable}{prop}{regionLattice}
  \label{prop:region-lattice}
  $(\region{N},\cup,\cap,\subseteq)$ is a bounded distributive lattice with top
  $\top = \bz{}^N$ and bottom $\bot = \emptyset$.
\end{restatable}
%
The set of regions $\region{N}$ is the target of our specification
language model, and of the code analysis. We further wrap this in a
labelled variant which provides information on multiplicity and approximation.

@@

We mark in our model the presence of modifiers such as
\texttt{readOnce} and \texttt{atMost}.  Approximation modifiers are interpreted as
injections into the $\mathsf{Approx}$ variant by $\interpApprox{-}$ in
\Cref{subfig:modifier-model}.  The $\textsf{Approx}$ type corresponds
to the presence or absence of the spatial approximation modifier, with
\textsf{exact} when there is no such modifier and \textsf{lower} and
\textsf{upper} for \term{atLeast} and \term{atMost} respectively. In a similar way,
multiplicity modifiers are interpreted as injections in the
$\mathsf{Mult}$ variant by $\interpMult{-}$, corresponding to
the presence or absence of the \term{readOnce} modifier as shown in
\Cref{subfig:modifier-model}.

@@

\begin{defn}
  \label{defn:subscript}
  Recall array subscript terms of the form $a(\bar{e})$ from
  \Cref{def:array-subs}. We interpret these terms with the partial
  interpretation $\interp{-}^{\mathit{aterm}} : \textit{array-term}
  \prightarrow{} \mathcal{P}(\bz{})^N$. The interpretation is defined when
  all indices are either constant or neighbourhood indices as defined in
  \Cref{def:neighbour-ix}.
%
  \begin{align*}
    \interp{a(\bar{e})}^{\mathit{aterm}} =
      \prod_{1 \leq i \leq N} \mathit{subscript}(\bar{e}_i) & &
  %
    \textit{subscript}(e) = \begin{cases}
      \{ c \} & e \equiv i \pm c \\
      \bz & e \; \mbox{is absolute}
    \end{cases}
  \end{align*}
\end{defn}
%
Note that this produces subscript schemes (a form of index scheme), but
these are not members of \textit{Interval} or $\textit{Region}_N$. In
the next section, we use this interpretation on array subscripts in
the static analysis producer preceding checking and inference.

@@

Recall that \Cref{lem:vector-intersect} states that the
intersection of index schemes is an index scheme. When
combined with \Cref{lem:interval-identities} (union and intersection
on holed intervals), we can extend this closure to \emph{interval} schemes.
Subscripts schemes do not enjoy the same closure propery as the intersection of
two singleton sets may be empty, disqualifying the result from 
being a subscript index. In the following sections, we do not take
intersection of subscript indices.

Union is not as flexible as
intersection. \Cref{lem:vector-union} does not form a closure for
indexing schemes in general. For this reason, we represent sets of
indexing schemes as unevaluated union terms, or \emph{union normal
  form}.  Since the model forms a distributive lattice as established
in \Cref{prop:region-lattice}, the intersections can be pushed inwards
using the distributive law leaving only unions at the outer-level. We
further exploit associativity of union to put unions into a cons-tree
making the unions effectively a non-empty list.  We do not, however,
attempt to canonicalise the representation. Later on
\Cref{lem:vector-union} and the index scheme subset are used to reduce the
size of union normal form.
